---
title: "Week 2"
subtitle: "Growth Curves"
summary: "Growth Curves"
author: "Josh Jackson"
date: '2019-09-05'
type: post
output:
  blogdown::html_page:
    toc: true
---



# Growth curves

## Between person models and cross sectional data

You already know this, but it gives us a chance to review regression

$$ {Y}_{i} = b_{0} + b_{1}X_{1} + b_{2}X_{2} + b_{3}X_{3}+... +\epsilon_{i} $$

$$ \hat{Y}_{i} = b_{0} + b_{1}X_{1} + b_{2}X_{2} + b_{3}X_{3}+... $$

Parameters are considered fixed where one regression value corresponds to everyone. I.e., that association between X1 and Y is the same for everyone. 

Each person has a Y, denoted by the subscript i, and each has a residual associated with them, also designated by i. 


```{r,message = FALSE, warning=FALSE}
library(readr)
example <- read_csv("~/Box/5165 Applied Longitudinal Data Analysis/ALDA/example copy.csv")
example$ID <- as.factor(example$ID)
# you can find the data on my github at: https://github.com/josh-jackson/ALDA/example%20copy.csv

```

Lets look at some data. These data examine older adults who came into a study up to six times over a six year period. Multiple cognitive, psychiatric and imaging assessments were done. Let's look at functional connectivity network called SMN7. 

```{r, warning=FALSE}
library(tidyverse)
library(ggplot2)
gg1 <- ggplot(example,
   aes(x = week, y = SMN7)) + geom_point() + stat_smooth(method = "lm")   
print(gg1)


```

What happens if we run a regression? 

```{r}
regression <- lm(SMN7 ~ week, data = example)
summary(regression)

```

## Within person models e.g., 2-level models

We saw this last time where we can think of everyone being run in a separate regression model. Here the lines connect the dots of the same people across time. 

```{r, warning=FALSE}
library(tidyverse)
gg2 <- ggplot(example,
   aes(x = week, y = SMN7, group = ID)) + geom_point() + stat_smooth(method = "lm", se = FALSE)   

gg3 <- gg2 +  stat_smooth(data = example, aes(x = week, y = SMN7, group=1, colour="#990000"), method = "lm", size = 3, se=FALSE) + theme(legend.position = "none")
print(gg3)


```

Each person has multiple assessments, so we need to distinguish between people and their assessments. In normal regression we wouldn't think about this as everyone datapoint is assumed to be independent. However, this is not the case here. Failing to distinguish would lead to violation of independence, an important assumption of the standard regression model. 

As seen in the graph above, what we have now is both individual level slopes as well as an average level slope. The average level slope is going to be the average of the individual level slopes, which will look like our average slope ignoring all dependencies. Same for the intercept. 

One way to do this is to run separate regressions for each person. Then we could just pool (or average) together where people start and how much they change to get the average intercept (starting value) and trajectory (how much people change). We will see later that this is a somewhat poor approach. 


```{r, echo = FALSE}
library(tidyverse)
library(broom)

set.seed(11)
ex.random <- example %>%
  dplyr::select(ID)%>% 
  distinct %>% 
  sample_n(10) 

example2 <-
  left_join(ex.random, example)  

```


```{r}
regressions <- example2 %>% 
  group_by(ID) %>% 
  do(tidy(lm(SMN7 ~ week, data = .)))

head(regressions)
```



In addition to the average intercept and the average trajectory there is also the amount of variation around each of these estimates. Do people tend to change the same? Are there individual differences in the initial assessment? 

This type of meaningful variation is lost when we have a between subjects only model that ignores the individual level. This variation will be called Random Effects (or variance estimates in SEM).

[Side note: note how some people do not have se estimates for their regression coefficients. The reason for this will impact our ability to fit longitudinal models later on.]

There is another important source of variation different from standard regression models. The within-subjects error that can be seen in the below graph. If we did not take people into account and just collapsed across people to get a between subjects assessment of change, this error would be confounded with individual differences in change. We will discuss this error more in depth later, but one way to think about our goal is to utilize our repeated assessments to make better predictions. A way to do that is to create additional buckets of explained variance, resulting in a smaller bucket of unexplained variance. 


```{r}
example3 <- example2 %>% 
  filter(ID == "67") 
```


```{r}
gg4 <-  ggplot(example3, aes(x = week, y = SMN7, group = ID)) +  geom_point() + stat_smooth(method = "lm")

gg4
```


### Thinking about random effects

### Random effects
Within subjects variability in either starting value or slope/trajectory is referenced in terms of random effects. How do we represent this in our equation? Easy, we just say that the typical regression parameters we have are not the same for everyone -- that they are random (in contrast to fixed).
 
In general, when would we want to use random effects? If there is some sort of selection (random or not) of many possible values of the predictor (e.g., stimuli are 3 random depression drugs, three semi random chosen levels of a drug). With longitudinal data this is (random) people. 

Side bar: Even in situations where these levels are not random (eg working with U.S. states) it is still useful to use MLM and we still call them random effects. To be consistent with language, random here can refer to as random from the population average, not randomly selected. When talking about "random effects" you can mean either of these definitions (and a few more). Luckily we can mostly ignore these semantic issues. 

What is necessary for modeling random effects? For longitudinal models, there needs to be multiple assessments per your grouping category (people, schools, neighborhoods, trials).  

We are assuming these random effects are sampled from some population and thus vary from group to group (or person to person). This means that your coefficients (like traditional regression coefficients) are estimates of some population parameter and thus have error associated with them. This error is not like like a standard residual, which represents error for your overall model. Nor is it like the standard error for a point estimate. Random effects can best be thought of as deviation of individual regression lines from the group regression line (though it technically is not this). 

To facilitate the multiple assessments per person we will now use both i and j subscripts. We will see that the random effects are part of the overall error term in the model. Counterintuitively, the main focus of these types of models will be the fixed effects, with less attention paid to the random effects. That said, the random effects are necessary to account for dependency in the data. One can think about these models as normal fixed effects regressions, with the random effects there to account for the longitudinal nature of the data. They are made up of a number of standard regression equations, each for a single individual. Doing so side steps the trouble of having correlated errors, and thus allows us to interpret our findings without concern. 

To facilitate adding random effects to our model it is helpful to think about two "levels" to our regression equation. We are going to put a regression equation within our regression equation. (Que Xzhibit joke). The first level will be the within-person model, in that it described how people differ across time. The second level will be the between person level. Note that these do not correspond to fixed or random effects. Instead they can be thought to model either within person differences or between person differences. Mastering thinking at these two levels will help make sense of these MLM models. 

### Empty model equation
Let's start with the most basic model and then expand from there.

Level 1 - within person
  $$ {Y}_{ij} = \beta_{0j}  +\varepsilon_{ij} $$
  
Note that we have multiple responses per individual j, noted with an i to refer to specific times.  

Also note that the intercept has a subscript. In typical regression it does not. This suggests that not everyone has the same intercept. 

The residuals at this level are thought of as measurement error OR as something that can be explained by time varying predictors. 
  
Level 2 - between person
  $$ {\beta}_{0j} = \gamma_{00} + U_{0j} $$
  
Level 2 takes the intercept (or other parameter) at level 1 and breaks it down into an equation for each individual, j. An overall group average (the gamma) and a residual term specific to deviation around the intercept (see below). 


And two variance components:
1. a random effect of the intercept
$$ {U}_{0j} \sim \mathcal{N}(0, \tau_{00}^{2})  $$
The subscript of the $U_{0j}$ refers to the number of the parameter where 0 is the intercept, 1 is the first regression coefficient, and so on. The second refers to the individual, j. So $U_{0j}$ refers to the intercept whereas $U_{1j}$ would refer to the random effect of the first regression coefficient. 

The $U_{0j}$ random effect is said to be normally distributed with a mean of zero and a variance of $\tau$

2. the residual error term
$$ {R}_{ij} \sim \mathcal{N}(0, \sigma^{2})  $$
Much like in normal regression there is an error term for all of the variation we cannot account for. What is unique here is that we took that normal variation and split it into two components. One that is attributable to variation around the intercept ${U}_{0j}$ and a catch all residual. 

Technically this is not a growth model, nor one that is inherently longitudinal. However, it does serve as a nice starting point to identify random effects.

###  Putting it together

  $$ {Y}_{ij} = \gamma_{00} + U_{0j}  + \varepsilon_{ij} $$
  
### Visualize what you are doing

Imagine the raw data plotted without knowing person j, how would  $\varepsilon_{i}$ be calculated? 

Now think about the data plotted again but with knowing each person has their own intercept. How would $\varepsilon_{ij}$ be calculated? 

Finally, how is $U_{0j}$ calculated? 
  

### ICC


If the ICC is greater than zero, we are breaking standard regression assumptions. 

$$\frac{U_{0j}}{U_{0j}+ \varepsilon_{ij}}$$


Is defined as % variation between over total variance. 

ICC can also be interpreted as the average (or expected) correlation within a nested group, in this case a person. On other words, the ICC is the correlation between any person's repeated measures (technically residuals). 


## Adding time
Here is the basic growth model where our predictor is a time variable

Level 1:
 
$$ {Y}_{ij} = \beta_{0j}  + \beta_{1j}X_{ij} + \varepsilon_{ij} $$
  
  Note how similar this looks like to a normal regression equation. Again, the differences are due to those pesky subscripts. Like before, think of this as a normal regression equation at the level of a person. Each person would have one of these equations with, in addition to a unique Y, X and residual, a unique $\beta_{0}$ and $\beta_{1}$. Look above to those individual regressions we did at the start of this section. 
  
 Level 2:  
$$ {\beta}_{0j} = \gamma_{00} + U_{0j}$$  
  
Level 2 takes the parameters at level 1 and decomposes them into a fixed component that reflects that average and then the individual deviations around that fixed effect. $U_{0j}$ is not error in the traditional sense. It describes how much variation there is around that parameter. Do some people start higher while some start lower, for example. 

$$ {\beta}_{1j} = \gamma_{10} $$  

The new level 2 term refers to the first predictor in the level 1 regression equation ie the slope. This slope is fixed in that the level 2 equation only has a gamma term and no U residual term. 
  
Putting it together: 
  $$ {Y}_{ij} = \gamma_{00} + \gamma_{10} (X_{1j})+ U_{0j}  + \varepsilon_{ij} $$
  
Note that in computing a single individuals Y, it depends on the two fixed effects, the Xj, and the random effect for the intercept. 


### What does this look like graphically? 
And how does this differ from the random intercept model? 

Can you draw out the sources of error? The random effects for each participant? The fixed effects? 


### Adding a random slope? 

What happens when we add a random slope? 
Level 1:
 
$$ {Y}_{ij} = \beta_{0j}  + \beta_{1j}X_{1j} + \varepsilon_{ij} $$
Level 2:  
$$ {\beta}_{0j} = \gamma_{00} + U_{0j}$$  

$$ {\beta}_{1j} = \gamma_{10} + U_{1j} $$  


Putting it together:
$$ {Y}_{ij} = \gamma_{00} + \gamma_{10}(X_{ij})+ U_{0j} + U_{1j}(X_{ij}) + \varepsilon_{ij} $$
  
  

Can think of a persons score divided up into a fixed component as well as the random component. 

These random effects are likely related to one another. For example, if someone starts high on a construct they are then less likely to increase across time. This negative correlation can be seen in the residual structure, where the random effects are again normally distributed with a mean of zero, but this time one must also consider covariance in addition to variance. 

$$ \begin{pmatrix} {U}_{0j} \\ {U}_{1j} \end{pmatrix}
\sim \mathcal{N} \begin{pmatrix} 
  0,     & \tau_{00}^{2} & \tau_{01}\\ 
  0, & \tau_{01} & \tau_{10}^{2}
\end{pmatrix} $$



Note that it is possible to have a different error structures, one where there is no relationship between the intercept and the slope, for example. We will discuss this more later in the semester. Right now just know that the default is to have correlated random effects. 

We also have the within subject variance term that accounts for deviations that are not accounted for by time variable and other level 1 predictors. 

$$ {R}_{ij} \sim \mathcal{N}(0, \sigma^{2})  $$

Note that it is possible to model these level 1 residuals with different structures. This specification implies that there is no correlation across an individuals residuals, once you account for level 1 predictors (ie growth trajectories). Having a specific level 1 autoregressive or other type of pattern is common in other treatments of longitudinal models (panel models) but is not necessary with growth models (but possible). 

This is the basic format of the growth model. It will be expanded later on by adding variables to the level 1 model and to the level 2 model. Adding to the level 1 model is only possible with repeated variables. 

Level 1 regression coefficients are added to the level 2 model. These coefficients are decomposed into a fixed effect, a random effect (possibly), and between person predictors. As with any regression model, each of these only have a single error term. 


## Individual level random effects

### Calculation of individual level random effects


Random effects are often thought in terms of variance components. We can see this if we think of individual level regressions for each person where we then have a mean and a variance for both the intercept or the slope. The greater the variance around the intercept and the slope means that not everyone starts at the same position and not everyone changes at the same rate. 

If you want to look at a specific person's random effect you can think of it as a deviation from the fixed effect where subject 6's intercept can be thought of as

$$ {\beta}_{06} = \gamma_{00} \pm U_{06}$$ 

e.g 2.2 = 3 - .8


### How are these random effects calculated? 
It isn't as straightforward as calculating a slope for each person and then using the difference between that slope and the average slope. Instead, the estimates are partially pooled towards the overall mean of the sample, the fixed effect. We do this to get a better estimate of the parameters, the same way that using regression to predict y-hat given an X is better than binning X and calculating y-hat. More information = better. 

Why not full pooling ie give everyone the same slope? Because it ignores individual differences in change. Often individual differences in (intraindividual) change is what we care about.  

The result is that the variance of the change trajectories (using MLM) will be smaller than the variance of the fitted linear models. Trajectories are "regressed" towards the average trajectory under the assumption that extreme scores are extreme because of (measurement) error, not that people are actually extreme. 

Can think about this in terms of creating an average for your intercept. Do you want the average to be the grand mean average, ignoring group? Do you want it to be the person average, ignoring that some people have more data points and thus are better assessed? No right answer, so maybe lets meet in the middle? This is sometimes called an empirical Bayes estimate. 


### Random effect decomposition

Think of the original total variance in a scatter plot of our DVs. Adding random effects takes that variance and trims it down. 

The intercept only MLM separates it into a level 1 variance (which at this stage is treated as error) and a level 2 random intercept variance. 

Creating a random slopes model takes the Level 1 residual variance and creates a new "pile" of explained or accounted for variance. 

We can then further explain the variance or reduce the pile by predictors at  level 1 and level 2. Our goal isn't necessarily to explain all of the variance but it is helpful to reduce the unexplained variance $\varepsilon_{ij}$ to improve model fit. 

### Random effects and residual (standard) assumptions

1. Joint normal distribution of random effects (Do we expect this to hold?)  
2. Normally distributed residual  
3. Constant variance over time  
4. Random effects $\pm U_{0j}$ and residual $\varepsilon_{ij}$ are uncorrelated and have a mean of zero  
 


## Data generating process (DGP)

Our standard assumption is that the DV comes from a data generating process that results in normal distributions. This does not mean that it needs to result in an observed normal distribution. Instead, the default of assuming an Gaussian DGP is practical: it is robust against violations and the alternatives are sometimes harder to justify. 

If you think you have a non-Gaussian DGP (like a Poisson or a negative binomial if you are using some sort of count data) you will need to use a different estimation technique. You can do this somewhat with the package we will be working with primarily, lme4. However, the BRMS package -- which uses Bayesian estimation -- has many more possibilities: geometric, log normal, weibull, exponential, gamma, Beta, hurdle Poisson/gamma/negative binomial, zero inflated beta/Poisson/negative binomial, cumulative. We will fit some of these later in the semester. Currently, however, assume we are working with  ${Y}_{ij} \sim \mathcal{N}(0, \sigma^{2})$


## Estimation

Maximum likelihood estimation. Uses a likelihood function that describes the probability of observing the sample data as a function of the parameters. Attempts to maximize the function through an iterative process. Because it is iterative, it might fail.  

There are fixed effects as well as random effects we need to count for. Maximum likelihood takes our assumptions about the model (normally distributed residuals, etc) and creates probability densities for each parameters. For example, based on certain fixed effects and sd of random effects, how likely is it that person x has a slope of z? The algorithm looks at the full sample to see how likely different parameters are, spits back the most likely, and gives you a number to show how likely they are (compared to others). This is akin to saying you rolled 10 dice, 5 came up as 2s. How likely is this dice fair? But instead of fair vs not fair it gives a likelihood to certain possibilities (e.g., a 2 comes up at 25%, 50% 75% rates). 


Restricted maximum likelihood (REML) vs Full Maximum likelihood (ML). Will give you similar parameters, the differences are in the standard errors. REML is similar to dividing by N - 1 for SE whereas ML is similar to dividing by N. 

Differences account for the fact that fixed effects are being estimated simultaneously with the variance parameters in ML. Estimates of the variance parameters assume that the fixed effects estimates are known and thus does not account for uncertainty in these estimates. 

REML accounts for uncertainty in the fixed effects before estimating residual variance. REML attempts to maximize the likelihood of the residuals whereas ML maximizes the sample data. REML can be thought of as an unbiased estimate of the residual variance. 

REML is good for small sample size both N and group. However, if you use REML you should be careful in testing fixed effects against each other (more down below). Deviances tests for fixed effects should be done with ML, but only random effects with REML. ML can also look at random effects too. 



## Testing significance (adapted from Ben Bolker)
4 Methods for testing single parameters
From worst to best:

1. Wald Z-tests. 

2. Wald t-tests

Easy to compute - test statistic over standard error However, they are asymptotic standard error approximations, assuming both that (1) the sampling distributions of the parameters are multivariate normal and that (2) the sampling distribution of the log-likelihood is (proportional to) χ2.

The above two are okay to do for single parameter estimates of fixed effects. But beware that  a) degrees of freedom calculations are not straightforward and b) the assumptions for random effects are be hard to meet. 

### Quick aside: P values are not included

Authors of the package we will be using first lme4 are not convinced of the utility of the general approach of testing with reference to an approximate null distribution. In general, it is not clear that the null distribution of the computed ratio of sums of squares is really an F distribution, for any choice of denominator degrees of freedom. While this is true for special cases that correspond to classical experimental designs (nested, split-plot, randomized block, etc.), it is apparently not true for more complex designs (unbalanced, GLMMs, temporal or spatial correlation, etc.).

tl;dr: it gets messy with more complex models.  

If you really want p values
```{r, eval = FALSE}
# library(lmerTest)
```


3. Likelihood ratio test (also called deviance test).  

4. Markov chain Monte Carlo (MCMC) or parametric bootstrap confidence intervals ( we will get to this later)


### Likelihood ratio test
Used for model comparisons (often multiparameter comparisons) and for tests of random effects. REML can only be used if model compared have the same fixed parts and only differ in random. Otherwise ML must be used. 

How much more likely the data is under a more complex model than under the simpler model (these models need to be nested to compare this).

Log Likelihood (LL) is derived from ML estimation. Logs are used because they are computationally simpler; logs of multiplications are reduced to adding the logs together. 

Larger the LL the better the fit. 

Deviance compares two LLs. Current model and a saturated model (that fits data perfectly). Asks how much worse the current model is to the best possible model. Deviance = -2[LL current - LL saturated]

LL saturated = 1 for MLMs (probability it will perfectly recapture data). log of 1 is 0. So this term drops out. Deviance = -2(LL current model). AKA -2logL or -2LL 

Can compare two models via subtraction, often referred to as a full and reduced model. Differences is distributed as a chi square with a df equal to how many "constraints" are included. Constraints can be thought of as forcing a parameter to be zero ie removing it.  

Comparing 2 models is called a likelihood ratio test. Need to have: 
1. same data
2. nested models (think of constraining a parameter to zero)

Why work with deviances and not just log likelihoods? Why -2? Why a ratio test when you subtract deviances? Maths. Working with deviances allows us to subtract two from one another, which is equivalent to taking the ratio of likelihoods. 

You can test in r using the same procedure we would to test different regression models.

```{r, eval = FALSE}
anova(mod.2, mod.2r)
```


### Likelihood tests for random effects
Not listed in the output because it is harder to do this with variances. Remember variances do not have values below zero and thus the distributions get a wonky quickly. Needs mixture distributions (Cannot be easily done with chi square, for example)

Can technically do anova comparisons for random effects, though that falls to many similar problems as trying to do a Wald test. 

The sampling distribution of variance estimates is in general strongly asymmetric: the standard error may be a poor characterization of the uncertainty. Thus the best way to handle is to do bootstrapped estimates. 

### AIC and BIC
Used when you want to compare non-nested data. Need to have the same data, however. 

AIC (Akaike’s Information Criterion) and the BIC (Bayesian Information Criterion) where “smaller is better.” This is the opposite of LL. As with the other types, these may give you wonky findings depending on some factors as they are related to LLs. 

AIC = 2(number of parameters) + (−2LL)
BIC = ln(n)(number of parameters) + (−2LL)

BIC penalizes models with more parameters more than AIC does. 

## Coefficient of determination equivalents
You want to get a model fit estimate. BIC and AIC are good to compare nested models but they aren't standardized and thus make comparison across non nested models difficult. 

With MLM models we cannot directly compute R2. Instead we will use pseudo R2. Pseudo R2 is similar to R2 in that it can be thought of as the correlation between your predicted and actual scores. For example, assume we have three waves of data. The intercept is 1, the slope is 2 and time is coded 0,1,2. The predicted scores are: 1, 3, 5. We would then correlate everyone's first, second and third wave scores with these predicted scores. This correlation squared is pseudo R2, telling us how much variance time explains in our DV. 

Yes, we typically think of this as a measure of variance explained divided by total variance. This is where things get tricky: should you include or exclude variation of different random-effects terms? These are error, but they are modeled in the sense that they are not unexplained. Is the effect size wanted after you are "controlling for" or do you want to talk about total variation. There are similarities here with regards to Eta and Partial Eta squared. 

The general idea is to be upfront about what you are comparing and what is included. Typically this is done with comparing models, much like a hierarchical regression. Taking the difference in variance between model 1 and model 2 and dividing it by model 1 makes it explicit what you are looking at and what you are including or not including. 

E.g,. residual variance in varying intercept model subtracted from growth model divided by intercept only model. This can tell you how much unexplained variance is explained by time. 

```{r, eval = FALSE}
(sigma(mod.1) - sigma(mod.2)) / sigma(mod.1)
```









