---
title: 'Week #7 & week #8'
author: ''
date: '2019-10-09'
slug: week-6
categories: []
tags: []
subtitle: 'SEM basics'
summary: 'SEM basics'
authors: []
lastmod: '2019-10-09T11:11:15-05:00'
featured: yes
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---


```{r setup}

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

```


# Longitudinal Structural Equation Modeling

SEM is the broader umbrella from the GLM. With it we are able to do two interesting this: 

1. Fit a latent measurement model (e.g., CFA)

2. Fit a structural model (e.g,. path analysis)

These two components allow us to address more difficult research questions involving but not limited to: multiple DVs, mediators, varying effects across time, unmeasured variables, constraints, and measurement invariance. 

Compared to MLM, and SEM approach to longitudinal data may be better suited to your research goals. We will see that at one level the two approaches can be equivalent. 


# SEM primer
In our studies some of the variables of interest will be measured and others will be latent or unobserved. Measured variables are the observed scores that are typically collected in a research study. That is, your variables in your dataset. This may include responses to survey items, reports of age or income, or performance on a recall task. We will use the general term "indicators" or "items" or "manifest variables" to refer to these measures variables. They will be indicated by boxes in a path diagram. 

Latent variables, in contrast, are not directly observed. A latent variable is something that we assume to exist but we cannot directly measure (see) it. They are typically postulated by theories or inferred from the statistical behavior of measured variables.  Sounds like psychological variables! Examples of latent variables include "depression," "job satisfaction," and "working memory capacity." Latent variables will be represented as a circle in path diagrams. 

For example, why does Sally like to go to parties, likes to talk a lot, and always tends to be a in a good mood? Maybe it is because her high levels of extraversion ( a latent variable that we cannot directly measure) is causing these tendencies. 

Key point: the variable/construct itself is not measurable, but the manifestations caused by the variable are measurable/observable. 

Interesting point: because variables are assumed to be causing indicators of the variable, SEM is sometimes referred to as causal modeling. (Also because in path models a directional relationship is hypothesized) Note that we cannot get any closer to causality than we can with regression. 

## Pretty pictures

Circles = latent variables

Boxes = observed indicator variables

two headed arrows = correlations/covariances/variances

single head arrows = regressions

triangle = means


```{r}
library(lavaan)
library(semPlot)

HolzingerSwineford1939 <- HolzingerSwineford1939

mod.1 <- 'visual =~ x1 + x2 + x3
textual =~ x4 + x5 + x6
speed =~ x7 + x8 + x9'

fit.1 <- cfa(mod.1, data=HolzingerSwineford1939,meanstructure = TRUE)
```


```{r}
semPaths(fit.1,  intercepts = TRUE)
```

```{r}
semPaths(fit.1, "std", intercepts = TRUE)
```


## Measurement model

The first step of an SEM model with latent variables is to define them. This is called specifying the measurement model. It is up to you to specify how you think the latent variable is created. It is then up to you to compare that measurement model against alternative measurement models to see if it meaningful. 

The key components are the a) factor loadings, b) residuals, and c) variance of latent variable. 

### Classical test theory interpretation 

How can we think of a latent construct:

Latent construct = what the indicators share in common

The indicators represent the sum of True Score variance + Item specific variance + Random error

The variance of the latent variable represents the amount of common information in the latent variable. If your indicators are haphazardly chosen then there will be low variance. We want to maximize variance here as variance suggests meaningful differences can be made between people. 

The residual errors (sometimes referred to as disturbances) represent the amount of information unique to each indicator. A combination of error and item-specific variance. 

The extent of the connection between the latent variable and the indicators is represented as a factor loading. 

### Generizability interpretation of latent variables
Same as above, but... 

True score variance can be thought of as consisting as a combination of 
1. Construct variance- this is the truest true score variance
2. Method variance- see Campbell and Fiske or sludge factor of Meehl. 
3. Occasion- important for longitudinal, aging, and cohort analyses--and for this class. 

For longitudinal models, occasion specific variance can lead to biased estimates. We want to separate the time specific variance from the overall construct variance. Or, we want to make sure that the time specific variance doesn't make it appear that a construct is changing when really it is not. 


### Formative indicators

These pretty pictures imply that the latent variables "cause" the indicators. This is the standard view and are referred to as reflexive indicators. However, there is another approach, formative indicators, were indicators "cause" the latent variable. Or, in other words, the latent variable doesn't actually exist. It is not real, only a combination of variables. An example of this is SES. SES does not 'exist' but is a socially constructed idea. Some people have argued that psychological constructs are of this kind, and that reflexive indicators are inappropriate. The arguments are mainly philosophical and thus is beyond the scope of our discussion. Suffice to say, that most of the time there are few analytic differences. You can think of this as similar to the factor analysis vs principal component analysis debates. 

### Measurement error

A major advantage is that each latent variable does not contain measurement error. It is as is if we measured our variable with an alpha = 1. 

What does that do? Well, ideally that gets us closer to the population model, which could yield higher R2 and parameter estimates.

How does this happen? It is a direct result of capturing what is shared among the indicators. The measurement error associated with each indicator is uncorrelated with the latent variable. 

Think about how this situation differs from creating a composite among variables. Think about how this differs from creating a factor score among variables within a simple factor analyses approach. How are all three different and similar? What does it mean if the error variances are correlated with one another? 

Remember however, that it is theoretically error free. The latent variable is not only filled with true score variance. Instead it could have method and occasion variance. Unless you have multiple methods and occasions it is hard to parse them apart. 


### Regarding means
SEM is also known as covariance structure analysis. You can do SEM using only variance-covariance matrices. These do not necessarily involve any direct information about their means. Means in SEM are optional. This is cool because you can technically reproduce the analyses of a paper if they give you a correlation matrix of study variables. 

Given we are interested in change across time, however, we will be interested in means. Latent variables by themselves do not have any inherent metric, it is up to us to choose the scale they are on. We can standardize them, use the original metric, and more! More later on how we define the mean of a latent variable.

## Path model

The path model component can be in addition to a measurement model or separate from them. You have already worked with path models as a simple regression is a path model, so is a standard mediation. You can make the path models more complex than these though, by specifying relationships among many variables. 

An example with no measurement model
```{r}
# generate data dataset:
X <- rnorm(100)
M <- rnorm(100)
Y <- rnorm(100) 
data <- data.frame(X, Y, M)

# Two regressions:
res1 <- lm(M ~ X, data = data)
res2 <- lm(Y ~ X + M, data = data)

# Plot mediation
semPaths(res1 + res2, "model", "est", intercepts = FALSE)

```


mediation model with measurement model
```{r}
mod.2 <- 'visual =~ x1 + x2 + x3
textual =~ x4 + x5 + x6
speed =~ x7 + x8 + x9

speed ~ textual + visual
textual ~ visual'

fit.2 <- cfa(mod.2, data=HolzingerSwineford1939)
semPaths(fit.2)
```



## Estimating an SEM model

Our goal when doing SEM is a creation of a model that specifies certain relationship among variables. This is done by creating a measurement or path model that we think is driven by the data generating process we are trying to study. In addition to setting the measurement model and paths we may want to put apriori constraining parameters (variances/covariances/regressions) to reflect how we think variables are related. 

E.g., Should these two variables be correlated or not? 

Then we use or ML algorithm to get our model implied covariances/means as close as possible to the observed covariances/means. 

Of note, SEM can handle any time of measured DV/IV or construct/indicators. If you have categorical indicators you can do SEM. However, it is hard to measure change using categorical indicators. But, categorical indicators are used for many latent variable models such as in measuring psychopathology. 

If you have a categorical construct you can also do SEM. Here it is called latent transition analysis (if you also had categorical indicators) or latent class / latent mixture modeling if you had continuous indicators (i am counting ordinal as continuous).


## Setting the scale and defining variables

We are trying to measure clouds. How can we do this given that they are always moving? Need to define the scale of a latent variable because there is no inherent scale of measurement. Largely irrelevant as to what scale is chosen just as centering or standardizing yield no substantive changes. Instead, scaling serves to establish a point of reference so as to interpret other parameters (much like the justifications for centering and standardizing). 

3 options: 

1. Fixed factor. Here you fix the variance of the latent variable to 1 (standardized).

2. Marker variable. Here you fix one factor loading to 1. All other loadings are relative to this loading. The variance of the latent variable can thus be anything. This is often the default of software programs. 

3. Effect coding. Here you constrain loading to average to 1. This will be helpful for us as we can then put the scale of measurement into our original metric. For longitudinal models this is helpful in terms of how to interpret the amount of change. 

## Identification

Note that we have multiple parameters we are trying to estimate. Paths (regression coefficients), means, variances (of manifest variables and latent variables). This makes estimating a little more tricky as you cannot have more unknowns than knowns. There are some tricks and rules of hand about what you can estimate or not. Typically however, it is suggested that you have 3 indicators per latent variable. 

If you are asking too much for a model you can also constrain parameters to be the same. For example, you may assume that all residual variances are the same. This assumption can be built into your model and reduces the number of parameters you have to estimate. 

More specifically, you need to compare the number of knows (variances and covariances) to the unknowns (model parameters). 

Foe example, a three indicator latent variable has 7 unknowns. 3 Loadings, 3 error variances and the variance of the latent variable

The covariance matrix has 6 data points. Thus we need to add in one more known, in this case a fixed factor or a marker variable. 

Knowns - unknowns = df. Note that df in this case df will not directly relate to sample size, so it is a little different than typical degree of freedom concepts. 

However, we will use this version of a df because it corresponds to how we will test the fit of these models. Specifically, the difference in dfs between models is distributed as a chi-square. 


### Types of identification

1. Just identified is where the number of knowns equal unknowns. Also known as saturated model. When you evaluate the fit of the model these will be perfect. So while these will estimate, we cannot examine whether or not our model is a good representation of the world, as we are simply recreating the observed covariance matrix (data). For some instances this may not be a problem, for others... 

2. Over identified is when you have more knowns than unknowns. This is good as we can fit a model that is more parsimonious than our data. Moreover, we can examine fit stats. 

3. Under identified is when you have problems and have more unknowns than knowns. this is because there is more than one solution available and the algorithm cannot decide e.g,. 2 + X = Y. If we add a constraint or a known value then it becomes manageable 2 + X = 12


### Fit Indices

1. residuals. Good to check.  

2. modification indices. Check to see if missing parameters that residuals may suggest you didn't include or should include. Can test with more advanced techniques. But eh... makes your models non-theoretical, could be over fitting, relying too much on sig tests...

3. chi-square. (Statistical fit) Implied versus observed data, tests to see if model are exact fit with data. But eh...too impacted by sample size

4. RMSEA or SRMR (Absolute fit). Does not compare to a null model. Judges distance from perfect fit. 

Above .10 poor fit
Below .08 acceptable

5. CFI, TFI (Relative fit models). Compares relative to a null model. Judges distance from the worse fit ie a null model. Null models have no covariance among observed and latent variables. 

range from 0-1. Indicate % of improvement from the null model to a saturated i.e. just identified model. 

Usually >.9 is okay. Some care about > .95

Minor changes to the model can improve fit. 

6. Check the model parameters. Are they wonky? Easy to get negative variances or correlations above 1. 



### Parcels
It is often necessary to simplify your model. One option to do so is with parcels where you combine indicators into a composite. This simplifies the model in that you have fewer parameters to fit. In addition to being a way to get a model identified, it also has benefits in terms of the assumptions of the indicator variables. 

To do so, you can combine items however you want into 3 or 4 groups or parcels, averaging them together. You may balance highly loading with less highly loading items (item to construct technique) or you may pair pos and negatively keyed items together. It is up to you. 

Some dislike it because you are aggregating without taking into account the association between the indicators; it is a blind procedure based on theory/assumptions rather than maths. ¯\_(ツ)_/¯



## Types of longitudinal models other than growth models (brief intro)


```{r}
long <- read.csv("~/Box/5165 Applied Longitudinal Data Analysis/SEM_workshop/longitudinal.csv")

summary(long)

```


### Longitudinal CFA

key concerns: 
1. Should the correlations be the same across time? 
2. Should the error variances be correlated? 
3. Are the loadings the same across time? (more on this later)



```{r}
long.cfa <- '
## define latent variables
Pos1 =~ PosAFF11 + PosAFF21 + PosAFF31
Pos2 =~ PosAFF12 + PosAFF22 + PosAFF32
Pos3 =~ PosAFF13 + PosAFF23 + PosAFF33
Neg1 =~ NegAFF11 + NegAFF21 + NegAFF31
Neg2 =~ NegAFF12 + NegAFF22 + NegAFF32
Neg3 =~ NegAFF13 + NegAFF23 + NegAFF33

## correlated residuals across time
PosAFF11 ~~ PosAFF12 + PosAFF13
PosAFF12 ~~ PosAFF13
PosAFF21 ~~ PosAFF22 + PosAFF23
PosAFF22 ~~ PosAFF23
PosAFF31 ~~ PosAFF32 + PosAFF33
PosAFF32 ~~ PosAFF33

NegAFF11 ~~ NegAFF12 + NegAFF13
NegAFF12 ~~ NegAFF13
NegAFF21 ~~ NegAFF22 + NegAFF23
NegAFF22 ~~ NegAFF23
NegAFF31 ~~ NegAFF32 + NegAFF33
NegAFF32 ~~ NegAFF33

'

fit.long.cfa <- cfa(long.cfa, data=long, std.lv=TRUE)

summary(fit.long.cfa, standardized=TRUE, fit.measures=TRUE)
```


```{r}
semPaths(fit.long.cfa)
```


### Longitudinal Path Model

key concerns: 
1. Should the regressions be the same across time? 
2. Should the error variances be correlated? 
3. Are the loadings the same across time? (more on this later)


```{r}
long.path <- '
## define latent variables
Pos1 =~ L1*PosAFF11 + L2*PosAFF21 + L3*PosAFF31
Pos2 =~ L1*PosAFF12 + L2*PosAFF22 + L3*PosAFF32
Pos3 =~ L1*PosAFF13 + L2*PosAFF23 + L3*PosAFF33
Neg1 =~ L4*NegAFF11 + L5*NegAFF21 + L6*NegAFF31
Neg2 =~ L4*NegAFF12 + L5*NegAFF22 + L6*NegAFF32
Neg3 =~ L4*NegAFF13 + L5*NegAFF23 + L6*NegAFF33

## free latent variances at later times (only set the scale once)
Pos2 ~~ NA*Pos2
Pos3 ~~ NA*Pos3
Neg2 ~~ NA*Neg2
Neg3 ~~ NA*Neg3
Pos1 ~~ Neg1
Pos2 ~~ Neg2
Pos3 ~~ Neg3

## directional regression paths
Pos2 ~ Pos1
Pos3 ~ Pos2
Neg2 ~ Neg1
Neg3 ~ Neg2

## correlated residuals across time
PosAFF11 ~~ PosAFF12 + PosAFF13
PosAFF12 ~~ PosAFF13
PosAFF21 ~~ PosAFF22 + PosAFF23
PosAFF22 ~~ PosAFF23
PosAFF31 ~~ PosAFF32 + PosAFF33
PosAFF32 ~~ PosAFF33

NegAFF11 ~~ NegAFF12 + NegAFF13
NegAFF12 ~~ NegAFF13
NegAFF21 ~~ NegAFF22 + NegAFF23
NegAFF22 ~~ NegAFF23
NegAFF31 ~~ NegAFF32 + NegAFF33
NegAFF32 ~~ NegAFF33
'

fit.long.path <- sem(long.path, data=long, std.lv=TRUE)

summary(fit.long.path, standardized=TRUE, fit.measures=TRUE)
```

```{r}
semPaths(fit.long.path, layout = "tree3")
## layout can also be done manually to get publications worthy plots

```


### Longitudinal Cross lagged panel model (clpm)

key concerns: 
1. Should the regressions (both cross lagged and autoregressive) be the same across time? 
2. Should the indicator error variances be correlated (within time or within construct)? 
3. Are the loadings the same across time? (more on this later)
4. Are the latent error variances the same or different? 
5. Are the latent error variances correlated the same or different across time? 
6. Are there more lagged effects? 


```{r}
long.cross <- '
## define latent variables
Pos1 =~ L1*PosAFF11 + L2*PosAFF21 + L3*PosAFF31
Pos2 =~ L1*PosAFF12 + L2*PosAFF22 + L3*PosAFF32
Pos3 =~ L1*PosAFF13 + L2*PosAFF23 + L3*PosAFF33
Neg1 =~ L4*NegAFF11 + L5*NegAFF21 + L6*NegAFF31
Neg2 =~ L4*NegAFF12 + L5*NegAFF22 + L6*NegAFF32
Neg3 =~ L4*NegAFF13 + L5*NegAFF23 + L6*NegAFF33

## free latent variances at later times (only set the scale once)
Pos2 ~~ NA*Pos2
Pos3 ~~ NA*Pos3
Neg2 ~~ NA*Neg2
Neg3 ~~ NA*Neg3

Pos1 ~~ Neg1
Pos2 ~~ Neg2
Pos3 ~~ Neg3

## directional regression paths
Pos2 ~ Pos1 + Neg1
Neg2 ~ Pos1 + Neg1
Pos3 ~ Pos2 + Neg2
Neg3 ~ Pos2 + Neg2

## correlated residuals across time
PosAFF11 ~~ PosAFF12 + PosAFF13
PosAFF12 ~~ PosAFF13
PosAFF21 ~~ PosAFF22 + PosAFF23
PosAFF22 ~~ PosAFF23
PosAFF31 ~~ PosAFF32 + PosAFF33
PosAFF32 ~~ PosAFF33

NegAFF11 ~~ NegAFF12 + NegAFF13
NegAFF12 ~~ NegAFF13
NegAFF21 ~~ NegAFF22 + NegAFF23
NegAFF22 ~~ NegAFF23
NegAFF31 ~~ NegAFF32 + NegAFF33
NegAFF32 ~~ NegAFF33
'

fit.long.cross <- sem(long.cross,data=long, std.lv=TRUE)

summary(fit.long.cross, standardized=TRUE, fit.measures=TRUE)

```

```{r}
semPaths(fit.long.cross)
```

```{r}
semPaths(fit.long.cross, layout = "tree3")
```




## Longitudinal mediation model

```{r}
#Do Self-Reported Social Experiences Mediate the Effect of Extraversion on Life Satisfaction and Happiness?
#number close friends
library(readr)
TSS_sub <- read_csv("~/Box/5165 Applied Longitudinal Data Analysis/Longitudinal/TSS_sub.csv")


scon.model6<-'
# define extraversion
bfie =~ a1bfi01 + a1bfi06r + a1bfi11 + a1bfi16 + a1bfi21r + a1bfi26 + a1bfi31r + a1bfi36

# correlated residuals
a1bfi11 ~~  a1bfi16
a1bfi06r ~~ a1bfi21r + a1bfi31r
a1bfi21r ~~ a1bfi31r + a1bfi01

#define social connection at 4 waves
hconnect=~h1clrel + h1satfr + h1sosat + h1ced05
jconnect=~j1clrel + j1satfr + j1sosat + j1ced05
kconnect=~k1clrel + k1satfr + k1sosat + k1ced05
mconnect=~m1clrel + m1satfr + m1sosat + m1ced05

#correlate residuals
h1clrel ~~ j1clrel + k1clrel + m1clrel
j1clrel ~~ k1clrel + m1clrel
k1clrel ~~ m1clrel
h1satfr ~~ j1satfr + k1satfr + m1satfr
j1satfr ~~ k1satfr + m1satfr
k1satfr ~~ m1satfr 
h1sosat ~~ j1sosat + k1sosat + m1sosat 
j1sosat ~~ k1sosat + m1sosat 
k1sosat ~~ m1sosat
h1ced05 ~~ j1ced05 + k1ced05 + m1ced05
j1ced05 ~~ k1ced05 + m1ced05
k1ced05 ~~ m1ced05

# same time covariances between extraversion, connection, satisfaction
bfie~~a1swls
hconnect ~~ h1swls
jconnect ~~ j1swls
kconnect ~~ k1swls

#regressions to calculate indiret effects
hconnect ~ a1*bfie + d1*a1swls 
jconnect ~ a2*bfie + d2*h1swls + m1*hconnect
kconnect ~ a3*bfie + d3*j1swls + m2*jconnect
mconnect ~ a4*bfie + d4*k1swls + m3*kconnect
h1swls ~ y1*a1swls + c1*bfie 
j1swls ~ y2*h1swls + c2*bfie + b1*hconnect
k1swls ~ y3*j1swls + c3*bfie + b2*jconnect 
m1swls ~ y4*k1swls + c4*bfie + b3*kconnect

#effects 
# extraversion -> connect (a)
# connect ->  swb (b)
# extraversion -> swb (c)
# auto-regressive connection (m)
# auto-regressive swb (y)

ind:= a1*b1*y3*y4 + a1*m1*b2*y4 + a1*m1*m2*b3 + a2*b2*y4 + a2*m2*b3 + a3*b3
total:= ind + c4 + c3*y4 + c2*y3*y4 + c1*y2*y3*y4
'
scon62 <- sem(scon.model6, data=TSS_sub, missing = "ml", fixed.x = FALSE)
summary(scon62, standardized=T, fit.measures=TRUE)

# use se = "bootstrap" in the fit function to get bootstrapped se

```


# Growth models

The above models are well suited to address between subjects questions, but does not get at a within subjects questions at all. There are some that can separate these effects, which we will turn to in a few weeks. However, we will start with the most basic model, the model you are familiar with, the growth model. 


The implementation of growth models in an SEM framework is very similar to the HLM framework. The major differences is how time is treated. Here, time variables must be the same for everyone in that each assessment point must have a particular variable name associated with it. That is, time is considered categorical in SEM, whereas in MLM it could be treated continuously. This requirement also makes a differences in how our data need to be structured. Whereas previously we had a time variable, now we indirectly include time into our model by specifying when variables were assessed. This has the consequence of necessitating a wide format, as opposed to the long format of MLM. 

Other than time, the idea behind the growth model is exactly the same.

## Coding time
One key these models is how you code time. Because we are working with qualitative time rather than continuous everyone has to have the same time structure. 


Let's use the long dataset from the previous Bayesian workshop, taken from chapter 4 of Singer and Willet. It is a three wave longitudinal study of adolescents. We are looking at alcohol use during the previous year, measured from 0 - 7. COA is a variable indicating the child's parents are alcoholics.

```{r}
alcohol1_pp <- read_csv("alcohol1_pp.csv")
head(alcohol1_pp)
```

```{r}
library(tidyverse)
alcohol.wide <- alcohol1_pp %>% 
  select(-X1, -age_14, -ccoa) %>% 
  pivot_wider(names_from = "age", 
              names_prefix = "alcuse_",
              values_from  = alcuse) 
head(alcohol.wide)
```



```{r, echo = FALSE}
model.1 <- '  i =~ 1*alcuse_14 + 1*alcuse_15 + 1*alcuse_16 
            s =~ 0*alcuse_14 + 1*alcuse_15 + 2*alcuse_16'
fit.1 <- growth(model.1, data=alcohol.wide)
summary(fit.1)
```


To fit a growth model within SEM we are going to create two latent variables: an intercept and a slope/trajectory. This is the same we did prior when we fit an MLM growth model with time as a predictor. 




```{r}

semPaths(fit.1)

```



How do we define these two latent factors? They are going to be made up of our repeated measures. We then add constraints on the loadings to force the latent variables to be interpreted how we want them to be interpreted. 

As before, the simplest way to think of the intercept is the initial value. While the slope indexes change. To obtain that, we will constrain the loadings to the intercept to be 1 for all repeated measures. This will obtain what is *constant* across time. 

But where is that constant located? Previously we could center or change our time variable to change the interpretation of the intercept. We do the same thing within SEM whereby we change the interpretation of intercept by changing how we define the slope parameter. As always, the intercept is defined as when X (or time in this case) equals zero. That is, the intercept is the mean of the DV when the predictor is 0, where we have time as the predictor. 

The slope loadings typically include 0, so as to make the intercept interpretable. However, it is up to you to define where zero goes and what the rest of the loadings are. A few rules: First, how you code the loadings represent the pattern of change you expect. 0,1,2 suggests a straight line that does not increase in speed. 0,1,5, suggests something different. Usually it is better to stick with linear change to start out with. 

Second, remember that this is just fancy regression. So the coefficients are interpreted as in regression whereas for a one unit change in the slope (time) corresponds to a coef change in your DV. Thus, you will change the magnitude of your slope parameter by choosing 0, .5, 1 versus 0,1,2 versus 0, 10, 20. As always, choose what makes sense to interpret. Changing the loadings (if they are just a multiple) will not change the substantive interpretation. 


## How do we do that in lavaan? 

```{r}
model.1 <- '  i =~ 1*alcuse_14 + 1*alcuse_15 + 1*alcuse_16 
            s =~ 0*alcuse_14 + 1*alcuse_15 + 2*alcuse_16'
fit.1 <- growth(model.1, data=alcohol.wide)
summary(fit.1)
```

## Compare with MLM
```{r}
library(lme4)
fit1.mlm <- lmer(alcuse ~  age + (age | id), data = alcohol1_pp)
summary(fit1.mlm)
```

centered MLM
```{r}
fit1.mlm.c <- lmer(alcuse ~  age_14 + (age_14 | id), data = alcohol1_pp)
summary(fit1.mlm.c)
```


```{r}
semPaths(fit.1, 'est')
```


```{r}
model.1 <- '  i =~ 1*alcuse_14 + 1*alcuse_15 + 1*alcuse_16 
            s =~ 0*alcuse_14 + .5*alcuse_15 + 1*alcuse_16'
fit.1 <- growth(model.1, data=alcohol.wide)
summary(fit.1)
```

## Interpretation of variances

Intercept will reflect our fixed effects whereas our random effects are indexed by variances of the latent variables. These are interpreted the same as before. More variance means that more people differ from the fixed effect. 

In MLM we assumed that the residual variances were equal. Here we can better model if that is the case. Right now we are able to fit separate variances for each repeated measure. These residual variances are interpreted as what is left over or what is unique at each time point that cannot be explained by the trajectory. 


### constraining slope to be fixed only

As with MLM we have options to handle the inclusion of random effects.

```{r}
model.2 <- '  i =~ 1*alcuse_14 + 1*alcuse_15 + 1*alcuse_16 
            s =~ 0*alcuse_14 + .5*alcuse_15 + 1*alcuse_16
               s ~~0*s'

fit.2 <- growth(model.2, data=alcohol.wide)
summary(fit.2)
           
```

### introducing covariates/predictors

```{r}
# a linear growth model with a time invariatnt covariate

model.3 <- '  i =~ 1*alcuse_14 + 1*alcuse_15 + 1*alcuse_16 
            s =~ 0*alcuse_14 + .5*alcuse_15 + 1*alcuse_16
               
# regressions
    
    i ~ male + coa
    s ~ male + coa
'
fit.3 <- growth(model.3, data = alcohol.wide)
summary(fit.3)
```


```{r}
library(lme4)
fit3.mlm <- lmer(alcuse ~  age_14 + male*age_14 + age_14*coa + (age_14 | id), data = alcohol1_pp)
summary(fit3.mlm)
```

### centered predictors

```{r}

alcohol.wide$coa.c <- scale(alcohol.wide$coa, center=TRUE, scale = FALSE)
alcohol.wide$male.c <- scale(alcohol.wide$male, center=TRUE, scale = FALSE)


model.4 <- '  i =~ 1*alcuse_14 + 1*alcuse_15 + 1*alcuse_16 
            s =~ 0*alcuse_14 + .5*alcuse_15 + 1*alcuse_16
               
# regressions
    
    i ~ male.c + coa.c
    s ~ male.c + coa.c
'
fit.4 <- growth(model.4, data = alcohol.wide)
summary(fit.4)
```

What is different what is the same? Notice that the main difference is the intercepts ie fixed effects. 


### introducing time varying covariates
```{r, eval = FALSE}
# a linear growth model with a time-varying covariate

model.5 <- '  i =~ 1*alcuse_14 + 1*alcuse_15 + 1*alcuse_16 
            s =~ 0*alcuse_14 + .5*alcuse_15 + 1*alcuse_16
               
# regressions
    
    i ~ male.c + coa.c
    s ~ male.c + coa.c
    
# time-varying covariates

    alcuse_14 ~ c1
    alcuse_15 ~ c2
    alcuse_16 ~ c3

'
fit.5 <- growth(model.5, data = alcohol.wide)


```


### latent basis model 


```{r}
model.6 <- '  i =~ 1*alcuse_14 + 1*alcuse_15 + 1*alcuse_16 
            s =~ 0*alcuse_14 + alcuse_15 + 1*alcuse_16
              
'

fit.6 <- growth(model.6, data = alcohol.wide)
summary(fit.6)
```

Does not change the fit of the model nor the implied means, but it can change your parameters by changing the time scaling. 



## Measurement Invariance (MI)

To meaningfully look at means, we need to have the means mean the same thing. In other words, without using the word mean, we need to make sure that the measurement of the construct is consistent across time. If it isn't, then what we may see as change actually reflect people responding to the indicators differently. For example, a common item on an extraversion scale is "Do you like to go to parties?" This is likely interpreted differently by a 20 year old compared to a 70 year old. This is due to what is normative, what parties look like that a typical 20 and 70 year old go to, etcetera. Another way to look at this is the item "2 x 3 = X, solve for X". The reasons that a 8 year old and a 18 year old get the item incorrect is likely for different reasons (ie knowledge vs not being careful). 

Maturation is the easiest way to see differences, but it also happens when you want to compare groups ie some anova design. This assumption is typically never critically examined.  

### types of MI
Configural (pattern). Typically always true with a decent measure of your construct. Can be tested through test statistics and eye-balling. Serves as default. 

Weak (metric/loading). Can be easily met. Not meeting this shows big problems, unless you are working with a really large dataset (where there is large power to find differences). 

Strong (Scalar/intercept). Need to meet this designation to run longitudinal models and look at means across time. 

Strict (residual/error variance). Not necessarily better than Strong, and does not need to be satisfied to use longitudinal models. Why might this not hold even if you are assessing the same construct? Hint: think of what residual variance is made up of. 

### Testing MI

configural (baseline)

```{r}
config <- '
## define latent variables
Pos1 =~ PosAFF11 + PosAFF21 + PosAFF31
Pos2 =~ PosAFF12 + PosAFF22 + PosAFF32
Pos3 =~ PosAFF13 + PosAFF23 + PosAFF33


## correlated residuals across time
PosAFF11 ~~ PosAFF12 + PosAFF13
PosAFF12 ~~ PosAFF13
PosAFF21 ~~ PosAFF22 + PosAFF23
PosAFF22 ~~ PosAFF23
PosAFF31 ~~ PosAFF32 + PosAFF33
PosAFF32 ~~ PosAFF33


'

config <- cfa(config, data=long, meanstructure=TRUE, std.lv=TRUE)

summary(config, standardized=TRUE, fit.measures=TRUE)
```

weak (constrain loadings) 

```{r}
weak <- '
## define latent variables
Pos1 =~ L1*PosAFF11 + L2*PosAFF21 + L3*PosAFF31
Pos2 =~ L1*PosAFF12 + L2*PosAFF22 + L3*PosAFF32
Pos3 =~ L1*PosAFF13 + L2*PosAFF23 + L3*PosAFF33


## free latent variances at later times (only set the scale once)
Pos2 ~~ NA*Pos2
Pos3 ~~ NA*Pos3

## correlated residuals across time
PosAFF11 ~~ PosAFF12 + PosAFF13
PosAFF12 ~~ PosAFF13
PosAFF21 ~~ PosAFF22 + PosAFF23
PosAFF22 ~~ PosAFF23
PosAFF31 ~~ PosAFF32 + PosAFF33
PosAFF32 ~~ PosAFF33


'

weak <- cfa(weak, data=long, meanstructure=TRUE, std.lv=TRUE)

summary(weak, standardized=TRUE, fit.measures=TRUE)


```


Strong (constrain loadings and intercepts)

```{r}
strong <- '
## define latent variables
Pos1 =~ L1*PosAFF11 + L2*PosAFF21 + L3*PosAFF31
Pos2 =~ L1*PosAFF12 + L2*PosAFF22 + L3*PosAFF32
Pos3 =~ L1*PosAFF13 + L2*PosAFF23 + L3*PosAFF33


## free latent variances at later times (only set the scale once)
Pos2 ~~ NA*Pos2
Pos3 ~~ NA*Pos3


## correlated residuals across time
PosAFF11 ~~ PosAFF12 + PosAFF13
PosAFF12 ~~ PosAFF13
PosAFF21 ~~ PosAFF22 + PosAFF23
PosAFF22 ~~ PosAFF23
PosAFF31 ~~ PosAFF32 + PosAFF33
PosAFF32 ~~ PosAFF33


## constrain intercepts across time
PosAFF11 ~ t1*1
PosAFF21 ~ t2*1
PosAFF31 ~ t3*1


PosAFF12 ~ t1*1
PosAFF22 ~ t2*1
PosAFF32 ~ t3*1


PosAFF13 ~ t1*1
PosAFF23 ~ t2*1
PosAFF33 ~ t3*1


## free latent means at later times (only set the scale once)
Pos2 ~ NA*1
Pos3 ~ NA*1'

strong <- cfa(strong, data=long, meanstructure=TRUE, std.lv=TRUE)

summary(strong, standardized=TRUE, fit.measures=TRUE)
```

Strict (loadings, intercept, residual variances)

```{r}
strict <- '
## define latent variables
Pos1 =~ L1*PosAFF11 + L2*PosAFF21 + L3*PosAFF31
Pos2 =~ L1*PosAFF12 + L2*PosAFF22 + L3*PosAFF32
Pos3 =~ L1*PosAFF13 + L2*PosAFF23 + L3*PosAFF33


## free latent variances at later times (only set the scale once)
Pos2 ~~ NA*Pos2
Pos3 ~~ NA*Pos3


## correlated residuals across time
PosAFF11 ~~ PosAFF12 + PosAFF13
PosAFF12 ~~ PosAFF13
PosAFF21 ~~ PosAFF22 + PosAFF23
PosAFF22 ~~ PosAFF23
PosAFF31 ~~ PosAFF32 + PosAFF33
PosAFF32 ~~ PosAFF33

## equality of residuals 
PosAFF11 ~~ r*PosAFF11 
PosAFF12 ~~ r*PosAFF12
PosAFF13 ~~ r*PosAFF13

PosAFF21 ~~ r*PosAFF21 
PosAFF22 ~~ r*PosAFF22
PosAFF23 ~~ r*PosAFF23

PosAFF31 ~~ r*PosAFF31 
PosAFF32 ~~ r*PosAFF32
PosAFF33 ~~ r*PosAFF33


## constrain intercepts across time
PosAFF11 ~ t1*1
PosAFF21 ~ t2*1
PosAFF31 ~ t3*1


PosAFF12 ~ t1*1
PosAFF22 ~ t2*1
PosAFF32 ~ t3*1


PosAFF13 ~ t1*1
PosAFF23 ~ t2*1
PosAFF33 ~ t3*1


## free latent means at later times (only set the scale once)
Pos2 ~ NA*1
Pos3 ~ NA*1'

strict <- cfa(strict, data=long, meanstructure=TRUE, std.lv=TRUE)

summary(strict, standardized=TRUE, fit.measures=TRUE)
```

Note that there are other types of MI that we could investigate, depending on what we are interested in. We could look at equality of latent means and variances, as well as regressions, if they were in the model. 


### Comparing the models
Usually done through chi-square difference test. But this is a very sensitive test, especially with larger samples. Better to look at changes in CFI. If delta is .01 or greater than maybe it shows misfit. 

```{r}
##Compare configural and weak model
anova(config, weak)
```


```{r}
##Compare weak and strong model
anova(weak, strong)

fitmeasures(weak)['cfi']
fitmeasures(strong)['cfi']
fitmeasures(strict)['cfi']

```



## Second order growth model

Repeated measures are latent. Why would we want to do this? 

```{r}
sec.order <- '
## define latent variables
Pos1 =~ NA*PosAFF11 + L1*PosAFF11 + L2*PosAFF21 + L3*PosAFF31
Pos2 =~ NA*PosAFF12 + L1*PosAFF12 + L2*PosAFF22 + L3*PosAFF32
Pos3 =~ NA*PosAFF13 + L1*PosAFF13 + L2*PosAFF23 + L3*PosAFF33

## intercepts
PosAFF11 ~ t1*1
PosAFF21 ~ t2*1
PosAFF31 ~ t3*1

PosAFF12 ~ t1*1
PosAFF22 ~ t2*1
PosAFF32 ~ t3*1

PosAFF13 ~ t1*1
PosAFF23 ~ t2*1
PosAFF33 ~ t3*1


## correlated residuals across time
PosAFF11 ~~ PosAFF12 + PosAFF13
PosAFF12 ~~ PosAFF13
PosAFF21 ~~ PosAFF22 + PosAFF23
PosAFF22 ~~ PosAFF23
PosAFF31 ~~ PosAFF32 + PosAFF33
PosAFF32 ~~ PosAFF33


## latent variable intercepts
Pos1 ~ 0*1
Pos2  ~ 0*1
Pos3  ~ 0*1

#model constraints for effect coding
## loadings must average to 1
L1 == 3 - L2 - L3
## means must average to 0
t1 == 0 - t2 - t3

i =~ 1*Pos1 + 1*Pos2 + 1*Pos3 
s =~ 0*Pos1 + 1*Pos2 + 2*Pos3 '


fit.sec.order <- growth(sec.order, data=long, missing = "ML")
summary(fit.sec.order, fit.measures=TRUE)

```


## Multiple groups


```{r}

group <- read_csv("~/Box/5165 Applied Longitudinal Data Analysis/Longitudinal/group.csv")

group1 <- 'Positive =~ P1 + P2 + P3
Negative =~ N1 + N2 + N3 

Positive ~~ 1*Positive
Negative ~~ 1*Negative

Positive ~~ Negative
'

fit.group.1 <- cfa(group1, data=group, std.lv=TRUE, group = "Grade")

summary(fit.group.1, standardized=TRUE, fit.measures=TRUE)

semPaths(fit.group.1,'est', panelGroups=TRUE)
```

Need to put in labels to fix the different parameters across models

```{r}


group2 <- 'Positive =~ c(L1,L1)*P1 + c(L2,L2)*P2 + c(L3,L3)*P3
Negative =~ N1 + N2 + N3 

Positive ~~ 1*Positive
Negative ~~ 1*Negative

Positive ~~ Negative
'

fit.group.2 <- cfa(group2, data=group, std.lv=TRUE, group = "Grade")

summary(fit.group.2, standardized=TRUE, fit.measures=TRUE)


```

We could now test whether constraining the groups makes the fit worse (similar to how we did it for measurement invariance)

```{r}
anova(fit.group.1,fit.group.2)
```

What if you wanted to make everything the same for a particular type of parameter? Lavaan has an easier answer:

```{r}
group3 <- 'Positive =~ P1 + P2 + P3
Negative =~ N1 + N2 + N3 

Positive ~~ 1*Positive
Negative ~~ 1*Negative

Positive ~~ Negative
'

fit.group.3 <- cfa(group3, data=group, std.lv=TRUE, group = "Grade", group.equal = c("loadings", "intercepts"))

summary(fit.group.3, standardized=TRUE, fit.measures=TRUE)

```

What is nice is that the output will label what parameters are constrained. Go ahead and look at the lavaan tutorial for more (residual variances, latent means, etc)

### measurement invariance revisited

Can you see how MI tests are just a special type of multiple group analysis? We could set up our data to have variables assessed at two different time points with the time demarcated by a group variable. The exact same findings from MI would be found with multiple group. It all depends on how you structure your data.  

### When to use 

For Longitudinal models you are faced with a choice: do you want to use group status to predict your slope, for example, or do you want to use multiple groups. The short answer is that it does not matter! The semi-longer answer is that including group as a predictor is simple while multiple groups analyses are more complex but also more flexible. Using one or the other depends on what your theory about where the group differences occur. If they occur in only the regression relationship indicating group mean differences then go with the easy option. If you think that the measurement model also may differ for your groups you have a more nuanced theory that necessitates multiple groups.

```{r, eval=FALSE}
Demo.growth$group <- cut(Demo.growth$x1.c, breaks=c(-Inf, 0, Inf), labels=c("low", "high"))

model.7 <- '
  # intercept and slope with fixed coefficients
    i =~ 1*t1 + 1*t2 + 1*t3 + 1*t4
    s =~ 0*t1 + 1*t2 + 2*t3 + 3*t4
  # regressions
    i ~ group
    s ~ group
'
fit.7 <- growth(model.7, data = Demo.growth)
summary(fit.7)

```




```{r, eval = FALSE}

model.8 <- '
  # intercept and slope with fixed coefficients
    i =~ 1*t1 + 1*t2 + 1*t3 + 1*t4
    s =~ 0*t1 + 1*t2 + 2*t3 + 3*t4

'
fit.8 <- growth(model.8, data = Demo.growth, group = "group")
summary(fit.8)

```


Notice how the slope and the intercept are larger for the high group, suggesting a difference between groups. 

```{r, eval = FALSE}
model.9 <- '
  # intercept and slope with fixed coefficients
    i =~ 1*t1 + 1*t2 + 1*t3 + 1*t4
    s =~ 0*t1 + 1*t2 + 2*t3 + 3*t4

'
fit.9 <- growth(model.9, data = Demo.growth, group = "group", group.equal = "means")
summary(fit.9)

```


Does constraining the latent means lead to a worse fit? 

```{r, eval = FALSE}
anova(fit.8, fit.9)
```


testing just the slope difference without constraining anything else. 

```{r, eval = FALSE}
model.10 <- '
  # intercept and slope with fixed coefficients
    i =~ 1*t1 + 1*t2 + 1*t3 + 1*t4
    s =~ 0*t1 + 1*t2 + 2*t3 + 3*t4

s ~  c(m1, m1)*1

'
fit.10 <- growth(model.10, data = Demo.growth, group = "group")
summary(fit.10)

```

```{r, eval = FALSE}
anova(fit.8, fit.10)
```

## Missing data
## multivariate models
### Planned missing data

## Power






